# Packages
```{r}
set.seed(42)

install.packages("tidyverse")
install.packages("rpart.plot")
install.packages("pROC")
install.packages("caret")
install.packages("randomForest")
install.packages("factoextra")
install.packages("ggplot2")

library(factoextra)
library(ggplot2)
library(tidyverse)
library(rpart)
library(rpart.plot)
library(pROC)
library(caret)
library(ipred) 
library(randomForest)  
library(gbm)           
```

# Features
id (Unique id for each patient)
age (Age of the patient in years)
origin (place of study)
sex (Male/Female)
cp chest pain type ([typical angina, atypical angina, non-anginal, asymptomatic])
trestbps resting blood pressure (resting blood pressure (in mm Hg on admission to the hospital))
chol (serum cholesterol in mg/dl)
fbs (if fasting blood sugar > 120 mg/dl)
restecg (resting electrocardiographic results)
-- Values: [normal, stt abnormality, lv hypertrophy]
thalach: maximum heart rate achieved
exang: exercise-induced angina (True/ False)
oldpeak: ST depression induced by exercise relative to rest
slope: the slope of the peak exercise ST segment
ca: number of major vessels (0-3) colored by fluoroscopy
thal: [normal; fixed defect; reversible defect]
num: the predicted attribute

#Data preprocseeing 
```{r}
# Heart dataset (UCI version mirrored on GitHub)
heart <- read.csv("https://raw.githubusercontent.com/sharmaroshan/Heart-UCI-Dataset/master/heart.csv")

# Quick check
glimpse(heart)
summary(heart)

# Ensure target is factor for classification
heart$target <- factor(heart$target, levels=c(0,1), labels=c("NoDisease","Disease"))

# NA handling 
heart <- heart %>% drop_na()
anyNA(heart)
```

# Visualization of the Dataset
```{r}
ggplot(heart, aes(target, fill = target)) +
  geom_bar() +
  scale_fill_manual(values = c("#4C9F70", "#D95F5F")) +
  theme_minimal(base_size = 14) +
  labs(title="Target Distribution", x="Condition", y="Count")

library(reshape2)

# Convert target to numeric (NoDisease = 0, Disease = 1)
heart_corr <- heart %>%
  mutate(target_num = ifelse(target == "Disease", 1, 0)) %>%
  select(where(is.numeric), target_num)

cor_mat <- cor(heart_corr)

# Visualisation
ggplot(melt(cor_mat), aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low="#2166AC", mid="white", high="#B2182B", midpoint=0) +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle=45, hjust=1),
        axis.title = element_blank()) +
  labs(title = "Correlation Heatmap (Including Target)")
```

# PCA 
```{r}
library(tidyverse)

numeric_features <- heart %>%
  select(-target) %>%
  select(where(is.numeric))

pca_scaled <- scale(numeric_features)

pca_model <- prcomp(pca_scaled, center = TRUE, scale. = TRUE)

pca_two <- as.data.frame(pca_model$x[, 1:2])
colnames(pca_two) <- c("PC1", "PC2")

heart_pca2 <- cbind(pca_two, target = heart$target)

glimpse(heart_pca2)
summary(heart_pca2)
```

# PCA Visualization
```{r}
numeric_features <- heart[, sapply(heart, is.numeric)]
numeric_features <- numeric_features[, !(colnames(numeric_features) == "target")]

pca_scaled <- scale(numeric_features)
pca_model <- prcomp(pca_scaled, center = TRUE, scale. = TRUE)

# Scree plot
fviz_eig(pca_model, addlabels = TRUE, barfill = "#2C7FB8", barcolor = "white") +
  ggtitle("Scree Plot: Explained Variance by PCA Components")

fviz_pca_var(
  pca_model,
  col.var = "black",     
  repel = TRUE           
) +
  ggtitle("PCA Variable Contribution Plot (PC1 vs PC2)")

fviz_pca_biplot(
  pca_model,
  geom.ind = "point",
  pointshape = 21,
  pointsize = 2,
  fill.ind = heart$target,
  col.ind = "black",
  col.var = "black",
  repel = TRUE
)
```

# Spilt the Datasest into 7:3
```{r}
set.seed(42)

train_index <- sample(1:nrow(heart_pca2), 0.7 * nrow(heart_pca2))
train <- heart_pca2[train_index, ]
test <- heart_pca2[-train_index, ]
```

# Tree (Baseline model)
```{r}
tree_fit <- rpart(target ~ ., data=train, method="class", minsplit=2, minbucket=1,cp=0)
rpart.plot(tree_fit, type=2, extra=104, fallen.leaves=TRUE, main = "Baseline Decision Tree")

pred_class <- predict(tree_fit, test, type="class")

# Probability prediction 
pred_prob <- predict(tree_fit, test, type="prob")[, "Disease"]

# Accuracy + F1 score
conf_matrix <- confusionMatrix(pred_class, test$target)
conf_matrix

# ROC Curve
roc_tree <- roc(test$target, pred_prob)
plot(roc_tree, col="#1F78B4", lwd=2, main = "ROC Curve – Decision Tree")
roc_auc <- auc(roc_tree)

# Cross-Entropy (Log Loss)
y_true <- ifelse(test$target == "Disease", 1, 0)
epsilon <- 1e-15  
pred_prob_safe <- pmin(pmax(pred_prob, epsilon), 1 - epsilon)

log_loss <- -mean(y_true * log(pred_prob_safe) + (1 - y_true) * log(1 - pred_prob_safe))
log_loss

# Store the results in a list
results <- list(
  Accuracy      = conf_matrix$overall["Accuracy"],
  F1_Score      = conf_matrix$byClass["F1"],
  ROC_AUC       = roc_auc,
  Cross_Entropy = log_loss
)

results
```

# Boundaries Graph for Tree
```{r}
library(ggplot2)
library(rpart)
library(rpart.plot)

tree_fit <- rpart(target ~ PC1 + PC2, data=train, method="class", minsplit=2, minbucket=1, cp=0)

# Create a grid of points covering the PCA space
x_range <- seq(min(test$PC1), max(test$PC1), length.out = 500)
y_range <- seq(min(test$PC2), max(test$PC2), length.out = 500)

# Create a meshgrid of these points
grid <- expand.grid(PC1 = x_range, PC2 = y_range)

# Predict the class labels for each point on the grid
grid$pred_class <- predict(tree_fit, newdata = grid, type = "class")

# Plot the decision boundaries
ggplot() +
  geom_tile(data = grid, aes(x = PC1, y = PC2, fill = pred_class), alpha = 0.3) +  
  geom_point(data = test, aes(x = PC1, y = PC2, color = target), size = 3) +  
  scale_color_manual(values = c("NoDisease" = "blue", "Disease" = "red")) +  
  scale_fill_manual(values = c("NoDisease" = "lightblue", "Disease" = "lightcoral")) +
  labs(title = "Decision Tree: PCA-based Decision Boundaries", x = "PC1", y = "PC2") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(legend.title = element_blank())
```

# Tree ROC (train vs test set)
```{r}
tree_fit <- rpart(target ~ ., data=train, method="class", minsplit=2, minbucket=1,cp=0)
rpart.plot(tree_fit, type=2, extra=104, fallen.leaves=TRUE, main = "Baseline Decision Tree")

pred_class <- predict(tree_fit, test, type="class")
pred_prob_test <- predict(tree_fit, test, type="prob")[, "Disease"]
pred_prob_train <- predict(tree_fit, train, type="prob")[, "Disease"]

# Accuracy and F1 score for test data
confusionMatrix(pred_class, test$target)

# ROC curve for test data
roc_tree_test <- roc(test$target, pred_prob_test)
plot(roc_tree_test, col="#1F78B4", lwd=2,
     main = "ROC Curve – Decision Tree")

# ROC curve for training data
roc_tree_train <- roc(train$target, pred_prob_train)
lines(roc_tree_train, col="#FF6347", lwd=2)

# AUC for both training and test data
auc_test <- auc(roc_tree_test)
auc_train <- auc(roc_tree_train)

# Add AUC values to the plot
legend("bottomright", 
       legend = c(paste("Test AUC = ", round(auc_test, 3)), paste("Train AUC = ", round(auc_train, 3))),
       col = c("#1F78B4", "#FF6347"), lwd = 2)

# Cross-Entropy (Log Loss) for test data
y_true <- ifelse(test$target == "Disease", 1, 0)
epsilon <- 1e-15  
pred_prob_safe <- pmin(pmax(pred_prob, epsilon), 1 - epsilon)

# Calculate Cross-Entropy (Log Loss)
log_loss <- -mean(y_true * log(pred_prob_safe) + (1 - y_true) * log(1 - pred_prob_safe))

log_loss
```

# Tree Result
```{r}
results <- list(
  Accuracy      = confusionMatrix(pred_class, test$target)$overall["Accuracy"],
  F1_Score      = confusionMatrix(pred_class, test$target)$byClass["F1"],
  ROC_AUC       = auc(roc_tree),
  Cross_Entropy = log_loss
)

results
```

# Bagging (Baseline)
```{r}
set.seed(42)

bag_basic <- bagging(
  target ~ .,
  data = train,
  nbagg = 25 
)

bag_pred_class <- predict(bag_basic, test, type="class")
bag_pred_prob <- predict(bag_basic, test, type="prob")[, "Disease"]

# Confusion matrix
bag_cm <- confusionMatrix(bag_pred_class, test$target)
bag_cm

# ROC
roc_bag <- roc(test$target, bag_pred_prob)
plot(
  roc_bag,
  col = "#33A02C", lwd = 2,
  main = "ROC Curve – Bagging Model"
)
auc(roc_bag)

# Cross-Entropy (Log Loss) 

y_true <- ifelse(test$target == "Disease", 1, 0)

epsilon <- 1e-15
bag_prob_safe <- pmin(pmax(bag_pred_prob, epsilon), 1 - epsilon)

bag_logloss <- -mean(
  y_true * log(bag_prob_safe) +
    (1 - y_true) * log(1 - bag_prob_safe)
)

# Results List (same format as tree)
bag_results <- list(
  Accuracy      = bag_cm$overall["Accuracy"],
  F1_Score      = bag_cm$byClass["F1"],
  ROC_AUC       = auc(roc_bag),
  Cross_Entropy = bag_logloss
)

bag_results
```


```{r}
set.seed(42)

bag_basic <- bagging(
  target ~ .,
  data = train,
  nbagg = 25
)

bag_pred_prob_test  <- predict(bag_basic, test,  type = "prob")[, "Disease"]
bag_pred_prob_train <- predict(bag_basic, train, type = "prob")[, "Disease"]

roc_bag_test  <- roc(test$target,  bag_pred_prob_test,  levels = c("NoDisease","Disease"), direction = "<")
roc_bag_train <- roc(train$target, bag_pred_prob_train, levels = c("NoDisease","Disease"), direction = "<")

plot(
  roc_bag_test,
  col = "#33A02C", lwd = 2,
  main = "ROC Curve – Bagging Model (Train vs Test)"
)
lines(roc_bag_train, col = "#FF6347", lwd = 2)

auc_test  <- auc(roc_bag_test)
auc_train <- auc(roc_bag_train)

legend(
  "bottomright",
  legend = c(
    paste("Test  AUC =", round(auc_test, 3)),
    paste("Train AUC =", round(auc_train, 3))
  ),
  col = c("#33A02C", "#FF6347"),
  lwd = 2
)
```

# Tuning the Bagging (no. of bag) 

# AUC vs no. of tree
```{r}
nbagg_values <- c(10, 25, 50, 100, 150, 200, 250, 300)

bag_results <- data.frame()

for (n in nbagg_values) {
  set.seed(42)
  model <- bagging(target ~ ., data=train, nbagg = n, coob = TRUE)
  
  pred_prob <- predict(model, test, type="prob")[, "Disease"]
  auc_val <- auc(roc(test$target, pred_prob))
  
  bag_results <- rbind(bag_results,
    data.frame(nbagg = n, AUC = auc_val)
  )
}

# Find the best AUC value and corresponding nbagg
best_row <- bag_results[which.max(bag_results$AUC), ]
best_nbagg <- best_row$nbagg
best_auc <- best_row$AUC

# Plot the AUC values
ggplot(bag_results, aes(x = nbagg, y = AUC)) +
  geom_line(color = "#1F78B4", size = 1.2) +
  geom_point(color = "#1F78B4", size = 3) +
  
  # Add a vertical line at the best AUC value
  geom_vline(xintercept = best_nbagg, linetype = "dashed", color = "red", size = 1) +
  
  # Highlight the best AUC point
  geom_point(aes(x = best_nbagg, y = best_auc), color = "red", size = 4) +

  
  scale_x_continuous(breaks = bag_results$nbagg) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Bagging Tuning: Number of Trees vs AUC",
    x = "Number of Trees (nbagg)",
    y = "AUC"
  )
```

# Accuracy VS no. of tree
```{r}
library(caret)
library(ggplot2)

nbagg_values <- c(10, 25, 50, 100, 150, 200)

bag_results <- data.frame()

for (n in nbagg_values) {
  set.seed(42)
  
  model <- bagging(target ~ ., data = train, nbagg = n, coob = TRUE)
  
  pred_prob <- predict(model, test, type = "prob")[, "Disease"]
  
  pred_class <- ifelse(pred_prob > 0.5, "Disease", "NoDisease")
  pred_class <- factor(pred_class, levels = c("NoDisease", "Disease"))
  
  acc_val <- confusionMatrix(pred_class, test$target)$overall["Accuracy"]
  
  bag_results <- rbind(
    bag_results,
    data.frame(nbagg = n, Accuracy = acc_val)
  )
}

# Identify best nbagg
best_row <- bag_results[which.max(bag_results$Accuracy), ]
best_nbagg <- best_row$nbagg
best_acc   <- best_row$Accuracy

bag_results

ggplot(bag_results, aes(x = nbagg, y = Accuracy)) +
  geom_line(color = "#1F78B4", size = 1.2) +
  geom_point(color = "#1F78B4", size = 3) +
  
  # Vertical line marking the best nbagg
  geom_vline(xintercept = best_nbagg, linetype = "dashed", color = "red", size = 1) +
  
  # Highlight the best point
  geom_point(aes(x = best_nbagg, y = best_acc), 
             color = "red", size = 4) +
  
  # Text label above the best point
  geom_text(aes(x = best_nbagg, y = best_acc, 
                label = paste0("Best: nbagg=", best_nbagg, "\nAcc=", round(best_acc, 3))),
            vjust = -1, color = "red", size = 5) +
  
  scale_x_continuous(breaks = seq(0, 200, 20)) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Bagging Tuning: Number of Trees vs Accuracy",
    x = "Number of Trees (nbagg)",
    y = "Accuracy"
  )
```

# This uses cross-entropy as preformance matrix vs no. of tree
```{r}
library(caret)
library(ggplot2)

nbagg_values <- c(10, 25, 50, 100, 150, 200)

bag_results <- data.frame()

for (n in nbagg_values) {
  set.seed(42)
  
  model <- bagging(target ~ ., data = train, nbagg = n, coob = TRUE)
  

  pred_prob <- predict(model, test, type = "prob")[, "Disease"]
  y_true <- ifelse(test$target == "Disease", 1, 0)
  
  # Cross entropy 
  epsilon <- 1e-15
  pred_prob_safe <- pmin(pmax(pred_prob, epsilon), 1 - epsilon)
  
  logloss_val <- -mean(
    y_true * log(pred_prob_safe) +
      (1 - y_true) * log(1 - pred_prob_safe)
  )
  
  bag_results <- rbind(
    bag_results,
    data.frame(nbagg = n, LogLoss = logloss_val)
  )
}

# Identify best nbagg (lowest log loss)
best_row <- bag_results[which.min(bag_results$LogLoss), ]
best_nbagg <- best_row$nbagg
best_logloss <- best_row$LogLoss

bag_results

# Plot with best value highlighted
ggplot(bag_results, aes(x = nbagg, y = LogLoss)) +
  geom_line(color = "#1F78B4", size = 1.2) +
  geom_point(color = "#1F78B4", size = 3) +

  # Vertical line marking the best nbagg
  geom_vline(xintercept = best_nbagg, linetype = "dashed", 
             color = "red", size = 1) +

  # Highlight the best point
  geom_point(aes(x = best_nbagg, y = best_logloss), 
             color = "red", size = 4) +

  # Text label
  geom_text(aes(x = best_nbagg, y = best_logloss,
                label = paste0("Best: nbagg=", best_nbagg,
                               "\nLogLoss=", round(best_logloss, 4))),
            vjust = -1, color = "red", size = 5) +

  scale_x_continuous(breaks = seq(0, 200, 20)) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Bagging Tuning: Number of Trees vs Cross-Entropy (Log Loss)",
    x = "Number of Trees (nbagg)",
    y = "Cross-Entropy (Log Loss)"
  )
```

# Try add minsplit=2, minbucket=1,cp=0 that same wtih previosu in tree
```{r}
set.seed(42)

bag_basic <- bagging(
  target ~ .,
  data = train,
  nbagg = 150,
  coob = TRUE, # This is Out-of-Bag (OOB) Error, since average 63% of data used to train in each bootstrap, remaining data is OOB, each tree can predict those data, it is already another form of cv
  control=rpart.control(method="class",minsplit=2, minbucket=1,cp=0)

)

bag_pred_class <- predict(bag_basic, test, type="class")
bag_pred_prob <- predict(bag_basic, test, type="prob")[, "Disease"]

# Accuracy + F1 Score

bag_cm <- confusionMatrix(bag_pred_class, test$target)
bag_cm

# ROC + AUC

roc_bag <- roc(test$target, bag_pred_prob)

plot(
  roc_bag,
  col = "#33A02C", lwd = 2,
  main = "ROC Curve – Bagging Model"
)

auc(roc_bag)

# Cross-Entropy (Log Loss) 

y_true <- ifelse(test$target == "Disease", 1, 0)

epsilon <- 1e-15
bag_prob_safe <- pmin(pmax(bag_pred_prob, epsilon), 1 - epsilon)

bag_logloss <- -mean(
  y_true * log(bag_prob_safe) +
    (1 - y_true) * log(1 - bag_prob_safe)
)

# Mean Squared Error

bag_mse <- mean((y_true - bag_prob_safe)^2)

# Results List (same format as tree)
bag_results <- list(
  Accuracy      = bag_cm$overall["Accuracy"],
  F1_Score      = bag_cm$byClass["F1"],
  ROC_AUC       = auc(roc_bag),
  Cross_Entropy = bag_logloss
)

bag_results
```

# Used the best parameter for prediction of Bagging
```{r}
set.seed(42)

bag_basic <- bagging(
  target ~ .,
  data = train,
  nbagg = 150,
  coob = TRUE, # This is Out-of-Bag (OOB) Error, since average 63% of data used to train in each bootstrap, remaining data is OOB, each tree can predict those data, it is already another form of cv
  #control=rpart.control(method="class",minsplit=2, minbucket=1,cp=0)
)

bag_pred_class <- predict(bag_basic, test, type="class")
bag_pred_prob <- predict(bag_basic, test, type="prob")[, "Disease"]

# Accuracy + F1 Score

bag_cm <- confusionMatrix(bag_pred_class, test$target)
bag_cm

# ROC + AUC

roc_bag <- roc(test$target, bag_pred_prob)

plot(
  roc_bag,
  col = "#33A02C", lwd = 2,
  main = "ROC Curve – Bagging Model"
)

auc(roc_bag)

# Cross-Entropy (Log Loss) 

y_true <- ifelse(test$target == "Disease", 1, 0)

epsilon <- 1e-15
bag_prob_safe <- pmin(pmax(bag_pred_prob, epsilon), 1 - epsilon)

bag_logloss <- -mean(
  y_true * log(bag_prob_safe) +
    (1 - y_true) * log(1 - bag_prob_safe)
)

# Mean Squared Error

bag_mse <- mean((y_true - bag_prob_safe)^2)

# Results List (same format as tree)
bag_results <- list(
  Accuracy      = bag_cm$overall["Accuracy"],
  F1_Score      = bag_cm$byClass["F1"],
  ROC_AUC       = auc(roc_bag),
  Cross_Entropy = bag_logloss
)

bag_results
```

#Bagging ROC (train vs test)
```{r}
set.seed(42)

bag_basic <- bagging(
  target ~ .,
  data = train,
  nbagg = 150,
  coob = TRUE
)

bag_pred_class <- predict(bag_basic, test, type="class")
bag_pred_prob_test <- predict(bag_basic, test, type="prob")[, "Disease"]

bag_pred_prob_train <- predict(bag_basic, train, type="prob")[, "Disease"]

bag_cm <- confusionMatrix(bag_pred_class, test$target)
bag_cm

roc_bag_test <- roc(test$target, bag_pred_prob_test)
plot(roc_bag_test, col="#33A02C", lwd=2, main = "ROC Curve – Bagging Model")

roc_bag_train <- roc(train$target, bag_pred_prob_train)
lines(roc_bag_train, col="#FF6347", lwd=2)

auc_test <- auc(roc_bag_test)
auc_train <- auc(roc_bag_train)

legend("bottomright", 
       legend = c(paste("Test AUC = ", round(auc_test, 3)), paste("Train AUC = ", round(auc_train, 3))),
       col = c("#33A02C", "#FF6347"), lwd = 2)

y_true_test <- ifelse(test$target == "Disease", 1, 0)
epsilon <- 1e-15
bag_prob_safe_test <- pmin(pmax(bag_pred_prob_test, epsilon), 1 - epsilon)

bag_logloss_test <- -mean(
  y_true_test * log(bag_prob_safe_test) +
  (1 - y_true_test) * log(1 - bag_prob_safe_test)
)

y_true_train <- ifelse(train$target == "Disease", 1, 0)
bag_prob_safe_train <- pmin(pmax(bag_pred_prob_train, epsilon), 1 - epsilon)

bag_logloss_train <- -mean(
  y_true_train * log(bag_prob_safe_train) +
  (1 - y_true_train) * log(1 - bag_prob_safe_train)
)
```

# Boundaries Graph: Bagging
```{r}
set.seed(42)
bag_basic <- bagging(
  target ~ PC1 + PC2,
  data = train,
  nbagg = 150,     
  coob = TRUE             
)

x_range <- seq(min(test$PC1), max(test$PC1), length.out = 500)
y_range <- seq(min(test$PC2), max(test$PC2), length.out = 500)

grid <- expand.grid(PC1 = x_range, PC2 = y_range)

grid$pred_class <- predict(bag_basic, newdata = grid, type = "class")

ggplot() +

  geom_tile(data = grid, aes(x = PC1, y = PC2, fill = pred_class), alpha = 0.3) +  
  geom_point(data = test, aes(x = PC1, y = PC2, color = target), size = 3) +  
  scale_color_manual(values = c("NoDisease" = "blue", "Disease" = "red")) +  
  scale_fill_manual(values = c("NoDisease" = "lightblue", "Disease" = "lightcoral")) +  
  labs(title = "Bagging Model: PCA-based Decision Boundaries", x = "PC1", y = "PC2") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(legend.title = element_blank())
```

# Boosting (Baseline)
```{r}
set.seed(42)

# Convert target to numeric 0/1 for gbm
train_gbm <- train
test_gbm  <- test

train_gbm$target <- ifelse(train$target == "Disease", 1, 0)
test_gbm$target  <- ifelse(test$target == "Disease", 1, 0)

boost_basic <- gbm(
  formula = target ~ .,
  data    = train_gbm,
  distribution = "bernoulli",  # Binary classification
  n.trees = 100,              # Number of boosting iterations 
  interaction.depth = 1,       # Tree depth 
  shrinkage = 0.001,            # Standard Learning rate
  n.minobsinnode = 10,         # Minimum observations per leaf
  bag.fraction = 0.5,          # Boosting
  cv.folds = 0,                #  CV  
)

boost_prob <- predict(boost_basic, test, n.trees = 100, type = "response")
boost_class <- ifelse(boost_prob > 0.5, "Disease", "NoDisease")
boost_class <- factor(boost_class, levels = c("NoDisease", "Disease"))

boost_cm <- confusionMatrix(boost_class, test$target)
boost_cm

roc_boost <- roc(test$target, boost_prob)

plot(
  roc_boost,
  col = "#FF7F00", lwd = 2,
  main = "ROC Curve – Boosting Model"
)

auc(roc_boost)
```

# Boosting (baseline) visualize AUC
```{r}
set.seed(42)

# Convert target to numeric 0/1 for gbm
train_gbm <- train
test_gbm  <- test

train_gbm$target <- ifelse(train$target == "Disease", 1, 0)
test_gbm$target  <- ifelse(test$target == "Disease", 1, 0)

boost_basic <- gbm(
  formula = target ~ .,
  data    = train_gbm,
  distribution = "bernoulli",
  n.trees = 100,
  interaction.depth = 1,
  shrinkage = 0.001,
  n.minobsinnode = 10,
  bag.fraction = 0.5,
  cv.folds = 0,
  verbose = FALSE
)

boost_prob_test  <- predict(boost_basic, test_gbm,  n.trees = 100, type = "response")
boost_prob_train <- predict(boost_basic, train_gbm, n.trees = 100, type = "response")

roc_boost_test  <- roc(test_gbm$target,  boost_prob_test,  direction = "<")
roc_boost_train <- roc(train_gbm$target, boost_prob_train, direction = "<")

plot(
  roc_boost_test,
  col = "#FF7F00", lwd = 2,
  main = "ROC Curve – Boosting Model (Train vs Test)"
)
lines(roc_boost_train, col = "#1F78B4", lwd = 2)

# --- AUCs + legend ---
auc_test  <- auc(roc_boost_test)
auc_train <- auc(roc_boost_train)

legend(
  "bottomright",
  legend = c(
    paste("Test  AUC =", round(auc_test, 3)),
    paste("Train AUC =", round(auc_train, 3))
  ),
  col = c("#FF7F00", "#1F78B4"),
  lwd = 2
)

# Print AUC values
auc_test
auc_train
```

#This part tunes the no. of tree
```{r}
library(caret)
library(pROC)
library(ggplot2)

ntree_values <- c(50, 100, 200, 300, 500, 800, 1000, 1500, 2000)

set.seed(42)

boost_tuning <- data.frame()

for (nt in ntree_values) {
  
  model <- gbm(
    target ~ .,
    data = train_gbm,
    distribution = "bernoulli",
    n.trees = nt,
    interaction.depth = 2,
    shrinkage = 0.01,
    n.minobsinnode = 10,
    bag.fraction = 1.0,
    verbose = FALSE
  )

  pred_prob <- predict(model, test_gbm, n.trees = nt, type = "response")
  pred_class <- ifelse(pred_prob > 0.5, 1, 0)
  pred_class <- factor(pred_class, levels = c(0, 1))
  true_factor <- factor(test_gbm$target, levels = c(0, 1))
  
  acc_val <- confusionMatrix(pred_class, true_factor)$overall["Accuracy"]
  
  # AUC
  auc_val <- auc(roc(test_gbm$target, pred_prob))
  
  # LogLoss
  epsilon <- 1e-15
  pred_prob_safe <- pmin(pmax(pred_prob, epsilon), 1 - epsilon)
  logloss_val <- -mean(
    test_gbm$target * log(pred_prob_safe) +
      (1 - test_gbm$target) * log(1 - pred_prob_safe)
  )
  
  # Append all metrics
  boost_tuning <- rbind(
    boost_tuning,
    data.frame(
      n.trees = nt,
      Accuracy = acc_val,
      AUC = auc_val,
      LogLoss = logloss_val
    )
  )
}

boost_tuning

# Calculate the best value for each metric
best_accuracy <- boost_tuning[which.max(boost_tuning$Accuracy), ]
best_auc <- boost_tuning[which.max(boost_tuning$AUC), ]
best_logloss <- boost_tuning[which.min(boost_tuning$LogLoss), ]

# Plot for AUC
ggplot(boost_tuning, aes(x = n.trees, y = AUC)) +
  geom_line(color="#FF7F00", size=1.2) +
  geom_point(color="#FF7F00", size=3) +
  geom_vline(xintercept = best_auc$n.trees, linetype = "dashed", color = "red", size = 1) +
  geom_point(aes(x = best_auc$n.trees, y = best_auc$AUC), color = "red", size = 4) +
  geom_text(aes(x = best_auc$n.trees, y = best_auc$AUC, label = paste0("Best: n.trees=", best_auc$n.trees, "\nAUC=", round(best_auc$AUC, 3))), vjust = -1, color = "red", size = 5) +
  theme_minimal(base_size=14) +
  labs(
    title="Boosting Tuning: n.trees vs AUC",
    x="Number of Trees",
    y="AUC"
  )

# Plot for Log Loss
ggplot(boost_tuning, aes(x = n.trees, y = LogLoss)) +
  geom_line(color="#1F78B4", size=1.2) +
  geom_point(color="#1F78B4", size=3) +
  geom_vline(xintercept = best_logloss$n.trees, linetype = "dashed", color = "red", size = 1) +
  geom_point(aes(x = best_logloss$n.trees, y = best_logloss$LogLoss), color = "red", size = 4) +
  geom_text(aes(x = best_logloss$n.trees, y = best_logloss$LogLoss, label = paste0("Best: n.trees=", best_logloss$n.trees, "\nLogLoss=", round(best_logloss$LogLoss, 4))), vjust = -1, color = "red", size = 5) +
  theme_minimal(base_size=14) +
  labs(
    title="Boosting Tuning: n.trees vs LogLoss",
    x="Number of Trees",
    y="Cross-Entropy Loss"
  )

# Plot for Accuracy
ggplot(boost_tuning, aes(x = n.trees, y = Accuracy)) +
  geom_line(color="#2C7FB8", size=1.2) +
  geom_point(color="#2C7FB8", size=3) +
  geom_vline(xintercept = best_accuracy$n.trees, linetype = "dashed", color = "red", size = 1) +
  geom_point(aes(x = best_accuracy$n.trees, y = best_accuracy$Accuracy), color = "red", size = 4) +
  geom_text(aes(x = best_accuracy$n.trees, y = best_accuracy$Accuracy, label = paste0("Best: n.trees=", best_accuracy$n.trees, "\nAcc=", round(best_accuracy$Accuracy, 3))), vjust = -1, color = "red", size = 5) +
  theme_minimal(base_size=14) +
  labs(
    title="Boosting Tuning: n.trees vs Accuracy",
    x="Number of Trees",
    y="Accuracy"
  )
```
The AUC graph shows the no. of tree do not really affect the result, so 800 is the best value in this case.

#This part is for tuning shrinkage and depth
```{r}
library(gbm)
library(pROC)
library(ggplot2)

set.seed(42)

# Shrinkage values and corresponding depth values
shrinkage_values <- c(0.1, 0.05, 0.01, 0.005)
depth_values <- list(
  "0.1"  = c(1, 2, 3, 4),
  "0.05" = c(1, 2, 3, 4),
  "0.01" = c(1, 2, 3, 4),
  "0.005"= c(1, 2, 3, 4)
)

grid_results <- data.frame()

for (sh in shrinkage_values) {
  for (d in depth_values[[as.character(sh)]]) {
    
    # Train the GBM model
    model <- gbm(
      target ~ .,
      data = train_gbm,
      distribution = "bernoulli",
      n.trees = 800,             
      interaction.depth = d,
      shrinkage = sh,
      n.minobsinnode = 10,
      bag.fraction = 1.0,
      verbose = FALSE
    )
    
    pred_prob <- predict(model, test_gbm, n.trees = 800, type = "response")
    pred_class <- ifelse(pred_prob > 0.5, 1, 0)  # Convert probabilities to class labels
    
    # AUC
    auc_val <- auc(roc(test_gbm$target, pred_prob))
    
    # LogLoss (safe version)
    epsilon <- 1e-15
    pred_prob_safe <- pmin(pmax(pred_prob, epsilon), 1 - epsilon)
    logloss_val <- -mean(
      test_gbm$target * log(pred_prob_safe) +
      (1 - test_gbm$target) * log(1 - pred_prob_safe)
    )
    
    # Accuracy
    accuracy_val <- mean(pred_class == test_gbm$target)
    

    grid_results <- rbind(
      grid_results,
      data.frame(
        shrinkage = sh,
        depth = d,
        AUC = auc_val,
        LogLoss = logloss_val,
        Accuracy = accuracy_val
      )
    )
  }
}

print(grid_results)

# Plotting the results
ggplot(grid_results, aes(x = factor(depth), y = factor(shrinkage), fill = LogLoss)) +
  geom_tile() +
  scale_fill_gradient(low = "#08519C", high = "#DEEBF7") +
  labs(
    title = "Boosting Grid Search (LogLoss)",
    x = "Interaction Depth",
    y = "Shrinkage"
  ) +
  theme_minimal(base_size = 14)

ggplot(grid_results, aes(x = factor(depth), y = factor(shrinkage), fill = Accuracy)) +
  geom_tile() +
  scale_fill_gradient(low = "#FF0000", high = "#00FF00") +
  labs(
    title = "Boosting Grid Search (Accuracy)",
    x = "Interaction Depth",
    y = "Shrinkage"
  ) +
  theme_minimal(base_size = 14)

ggplot(grid_results, aes(x = factor(depth), y = factor(shrinkage), fill = AUC)) +
  geom_tile() +
  scale_fill_gradient(low = "#FFF3B0", high = "#D62828") +  # pale yellow -> deep red
  labs(
    title = "Boosting Grid Search (AUC)",
    x = "Interaction Depth",
    y = "Shrinkage"
  ) +
  theme_minimal(base_size = 14)
```

#Final Tuning: Boosting
```{r}
boost_basic <- gbm(
  formula = target ~ .,
  data = train_gbm,
  distribution = "bernoulli",  # Binary classification
  n.trees = 800,              # Number of boosting iterations 
  interaction.depth = 2,       # Tree depth 
  shrinkage = 0.01,            # Standard Learning rate
  n.minobsinnode = 1,         # Minimum observations per leaf
  bag.fraction = 1,          # Boosting
)

boost_prob_test <- predict(boost_basic, test, n.trees = 800, type = "response")
boost_class_test <- ifelse(boost_prob_test > 0.5, "Disease", "NoDisease")
boost_class_test <- factor(boost_class_test, levels = c("NoDisease", "Disease"))

# Confusion Matrix for test data
boost_cm <- confusionMatrix(boost_class_test, test$target)
boost_cm

# ROC for test data
roc_boost_test <- roc(test$target, boost_prob_test)

plot(
  roc_boost_test,
  col = "#FF7F00", lwd = 2,
  main = "ROC Curve – Boosting Model"
)

auc(roc_boost_test)

boost_prob_train <- predict(boost_basic, train_gbm, n.trees = 800, type = "response")

boost_class_train <- ifelse(boost_prob_train > 0.5, "Disease", "NoDisease")
boost_class_train <- factor(boost_class_train, levels = c("NoDisease", "Disease"))

# ROC for train data
roc_boost_train <- roc(train_gbm$target, boost_prob_train)
lines(roc_boost_train, col = "#1F78B4", lwd = 2)
auc_train <- auc(roc_boost_train)

# Display both AUC values
legend("bottomright", 
       legend = c(paste("Test AUC = ", round(auc(roc_boost_test), 3)),
                  paste("Train AUC = ", round(auc_train, 3))),
       col = c("#FF7F00", "#1F78B4"), lwd = 2)

# Cross-Entropy (Log Loss) 
y_true_test <- ifelse(test$target == "Disease", 1, 0)

epsilon <- 1e-15
boost_prob_safe_test <- pmin(pmax(boost_prob_test, epsilon), 1 - epsilon)

boost_logloss_test <- -mean(
  y_true_test * log(boost_prob_safe_test) +
  (1 - y_true_test) * log(1 - boost_prob_safe_test)
)

# Mean Squared Error for test data
boost_mse_test <- mean((y_true_test - boost_prob_safe_test)^2)

# Cross-Entropy (Log Loss) for train data
y_true_train <- ifelse(train_gbm$target == "Disease", 1, 0)

boost_prob_safe_train <- pmin(pmax(boost_prob_train, epsilon), 1 - epsilon)

boost_logloss_train <- -mean(
  y_true_train * log(boost_prob_safe_train) +
  (1 - y_true_train) * log(1 - boost_prob_safe_train)
)

# Results List 
boost_results <- list(
  Accuracy      = boost_cm$overall["Accuracy"],
  F1_Score      = boost_cm$byClass["F1"],
  ROC_AUC_Test  = auc(roc_boost_test),
  Cross_Entropy_Test = boost_logloss_test
)

boost_results
```

# Boundaries Graph gbm
```{r}
set.seed(42)
boost_basic <- gbm(
  formula = target ~ .,   
  data = train_gbm,           
  distribution = "bernoulli", 
  n.trees = 800,              
  interaction.depth = 2,    
  shrinkage = 0.01,             
  n.minobsinnode = 1,       
  bag.fraction = 1          
)

x_range <- seq(min(test$PC1), max(test$PC1), length.out = 500)
y_range <- seq(min(test$PC2), max(test$PC2), length.out = 500)

grid <- expand.grid(PC1 = x_range, PC2 = y_range)

grid$pred_prob <- predict(boost_basic, newdata = grid, n.trees = 800, type = "response") 
grid$pred_class <- ifelse(grid$pred_prob > 0.5, "Disease", "NoDisease")

# Step 5: Plot the decision boundaries for the GBM model
ggplot() +
  geom_tile(data = grid, aes(x = PC1, y = PC2, fill = pred_class), alpha = 0.3) +  
  geom_point(data = test, aes(x = PC1, y = PC2, color = factor(target)), size = 3) + 
  scale_color_manual(values = c("NoDisease" = "blue", "Disease" = "red")) +  
  scale_fill_manual(values = c("NoDisease" = "lightblue", "Disease" = "lightcoral")) +  
  labs(title = "GBM Model: PCA-based Decision Boundaries", x = "PC1", y = "PC2") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(legend.title = element_blank())
```

# Final result for three models
```{r}
library(dplyr)

final_table <- data.frame(
  Model = c("Decision Tree", "Tuned Bagging", "Tuned Boosting"),
  
  Accuracy = c(
    results$Accuracy,
    bag_results$Accuracy,
    boost_results$Accuracy
  ),
  
  F1_Score = c(
    results$F1_Score,
    bag_results$F1_Score,
    boost_results$F1_Score
  ),
  
  ROC_AUC = c(
    results$ROC_AUC,
    bag_results$ROC_AUC,
    boost_results$ROC_AUC
  ),
  
  Cross_Entropy = c(
    results$Cross_Entropy,
    bag_results$Cross_Entropy,
    boost_results$Cross_Entropy
  )
)

final_table
```

# ROC graph for three models
```{r}
plot(roc_tree, 
     col = "#1F78B4", lwd = 2, 
     main = "ROC Comparison: Tree vs Bagging vs Boosting")

plot(roc_bag, 
     col = "#33A02C", lwd = 2, 
     add = TRUE)

plot(roc_boost, 
     col = "#FF7F00", lwd = 2, 
     add = TRUE)

legend(
  "bottomright",
  legend = c(
    paste0("Decision Tree (AUC = ", round(results$ROC_AUC,3), ")"),
    paste0("Bagging (AUC = ", round(bag_results$ROC_AUC,3), ")"),
    paste0("Boosting (AUC = ", round(boost_results$ROC_AUC,3), ")")
  ),
  col = c("#1F78B4", "#33A02C", "#FF7F00"),
  lwd = 2,
  cex = 0.8
)
```

# Hence PCA compresses 13 heart-disease features into 2 principal components, the decision boundary becomes low-variance and highly linear. Bagging reduces variance, but variance is already low in PCA space, so bagging does not bring additional performance gains. Boosting, on the other hand, reduces bias through sequential weak learners, which is more beneficial under PCA compression.